Adaptive Shrinkage and False Discovery Rates by Laplace Approximation

### 2014-02-24 paper/Rcode/conservativepm.rmd
added uc talk

```{r}
#Plot pi0 from each method
  get_pi0.fdrtool = function(f){f$param[3]}
  get_pi0.locfdr = function(f){f$fp0[1,3]}
  get_pi0.mixfdr = function(f){f$pi[1]}

  plot_pi0 = function(sims){
    res=list()
    for(i in 1:length(sims)){
      pi0=sims[[i]]$pi0  
      pi0_ash.n=unlist(lapply(sims[[i]]$betahat.ash.n,get_pi0))
      pi0_ash.u=unlist(lapply(sims[[i]]$betahat.ash.u,get_pi0)) 
      pi0_fdrtool = unlist(lapply(sims[[i]]$betahat.fdrtool,get_pi0.fdrtool))
      pi0_locfdr=unlist(lapply(sims[[i]]$betahat.locfdr,get_pi0.locfdr))
      pi0_mixfdr = unlist(lapply(sims[[i]]$betahat.mixfdr,get_pi0.mixfdr))
      pi0_qval = unlist(lapply(sims[[i]]$betahat.qval,"[[","pi0"))

      res[[i]] = data.frame(Scenario=i,pi0=pi0,qvalue=pi0_qval,mixfdr=pi0_mixfdr, locfdr=pi0_locfdr, fdrtool=pi0_fdrtool,ash.nullbiased=pi0_ash.n,ash.uniform=pi0_ash.u)
    }
    cbbPalette <- c("#56B4E9", "#009E73", "#F0E442", "#0072B2", "#E69F00", "#000000", "#D55E00", "#CC79A7")
    labels = c('mixfdr','qvalue','locfdr','fdrtool','ash (null-biased)','ash (mle)')
    breaks = c("mixfdr","qvalue","locfdr","fdrtool","ash.nullbiased","ash.uniform")

    require(reshape2)
    res.melt = melt(res, id.vars=c("pi0","Scenario"),variable.name="Method")
    res.melt = res.melt[sample(1:nrow(res.melt)),]  
    res.melt$Scenario = as.factor(res.melt$Scenario)
    levels(res.melt$Scenario) = c("Scenario 1a","Scenario 1b","Scenario 2")
    p=ggplot(data=res.melt,aes(pi0,value,colour=Method)) +geom_point(shape=1) +
        facet_grid(. ~ Scenario) +
      geom_abline(colour = "black") +
        xlab("True pi0") +
        ylab("Estimated pi0")
    print(p +scale_y_continuous(limits=c(0,1)) +
          scale_x_continuous(limits=c(0,1)) +
          scale_colour_manual(values=cbbPalette,breaks=breaks,labels=labels) +
          coord_equal(ratio=1))

@@

  png("figures/estpi0_sim1sim2.png",height=160,width=540)
    plot_pi0(list(simres1,simres1a,simres2))
  dev.off()

@@

```{r}
mean_quant=function(x, mult = 1){
      x <- na.omit(x)
      sd <- mult * sqrt(var(x))
      mean <- mean(x)
      data.frame(y = median(x), ymin = quantile(x,0.25) , ymax = quantile(x,0.75))
}

#ptype indicates what type of plot to do
#maxlfsr controls maximum x axis value
#maxy controls maximum y axis value
  plot_lfsr=function(sims,maxlfsr=0.1,ptype=c("lfsr","lfsra","lfdr"),maxy=1){
    ptype=match.arg(ptype)
    xlabtype = ifelse(ptype=="lfdr","lfdr","lfsr")
    res=list()
    for(i in 1:length(sims)){
      lfsr.ash.n= unlist(lapply(sims[[i]]$betahat.ash.n,"[[",ptype))
      lfsr.ash.u= unlist(lapply(sims[[i]]$betahat.ash.u,"[[",ptype))
      if(ptype=="lfdr"){
        lfsr.ash.true= unlist(lapply(sims[[i]]$betahat.ash.true,"[[","lfdr"))
        lfdr.mixfdr = unlist(lapply(sims[[i]]$betahat.mixfdr,"[[","fdr"))
        lfdr.locfdr = unlist(lapply(sims[[i]]$betahat.locfdr,"[[","fdr"))
      } else {
         lfsr.ash.true= unlist(lapply(sims[[i]]$betahat.ash.true,"[[","lfsr"))
      }

      subset = lfsr.ash.true<maxlfsr

      if(length(subset)>0){  
        res[[i]] = data.frame(Scenario=i,ash.nullbiased=lfsr.ash.n[subset],
                      ash.uniform=lfsr.ash.u[subset],
                      Bayes=0.1*maxlfsr*findInterval(lfsr.ash.true[subset],seq(0,maxlfsr,length=11))-0.05*maxlfsr)
        if(ptype=="lfdr"){
          res[[i]]=data.frame(res[[i]],mixfdr = lfdr.mixfdr[subset])
        }
      }
    }

    require(reshape2)

    res.melt = melt(res, id.vars=c("Bayes","Scenario"),variable.name="Method")

    cbbPalette <- c("#000000","#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
    labels = c('ash (null-biased)','ash (mle)')
    breaks = c("ash.nullbiased","ash.uniform")
    if(ptype=="lfdr"){
      labels = c('mixfdr', labels)
      breaks = c("mixfdr", breaks)
    }

    p=ggplot(data=res.melt,aes(Bayes,value,colour=Method)) + 
        facet_grid(. ~ Scenario) +
        #scale_fill_manual(values=cbbPalette) +
        #scale_colour_manual(values=cbbPalette) +
        geom_point(size=1,alpha=0.1) + 
        stat_smooth(se=FALSE,size=2) +
        stat_summary(fun.data="mean_quant", geom="ribbon", alpha=0.25) +
        geom_abline(colour = "black") +
        xlab(paste0(xlabtype," (Bayes)")) +
        ylab(paste0(ptype," (Method)"))

    print(p +scale_y_continuous(limits=c(0,maxy)) +
          scale_x_continuous(limits=c(0,maxlfsr)) +
            scale_colour_manual(values=cbbPalette,breaks=breaks,labels=labels))         
  }

```{r}
#plots estimated lfdr and lfsr against truth.
#ptype indicates whether to use lfsr or lfsra
#maxx controls maximum x axis value
#maxy controls maximum y axis value
plot_lfsdr=function(sims,maxx=0.1,ptype=c("lfsr","lfsra"),maxy=1){
    ptype=match.arg(ptype)
    res=list()
    res.lfsr=list()
    res.lfdr=list()

    for(i in 1:length(sims)){
      lfsr.ash.true= unlist(lapply(sims[[i]]$betahat.ash.true,"[[","lfsr"))
      lfsr.ash.n= unlist(lapply(sims[[i]]$betahat.ash.n,"[[",ptype))
      lfsr.ash.u= unlist(lapply(sims[[i]]$betahat.ash.u,"[[",ptype))

      lfdr.ash.n= unlist(lapply(sims[[i]]$betahat.ash.n,"[[","lfdr"))
      lfdr.ash.u= unlist(lapply(sims[[i]]$betahat.ash.u,"[[","lfdr"))
      lfdr.ash.true= unlist(lapply(sims[[i]]$betahat.ash.true,"[[","lfdr"))
      lfdr.mixfdr = unlist(lapply(sims[[i]]$betahat.mixfdr,"[[","fdr"))
      lfdr.locfdr = unlist(lapply(sims[[i]]$betahat.locfdr,"[[","fdr"))

      subset = lfsr.ash.true<maxx

      res.lfsr[[i]] = data.frame(Scenario=i,Measure='lfsr', ash.nullbiased=lfsr.ash.n[subset], ash.uniform=lfsr.ash.u[subset],Bayes=0.1*maxx*findInterval(lfsr.ash.true[subset],seq(0,maxx,length=11))-0.05*maxx,mixfdr = NA)

      subset = lfdr.ash.true<maxx
      res.lfdr[[i]] = data.frame(Scenario=i,Measure='lfdr', ash.nullbiased=lfdr.ash.n[subset], ash.uniform=lfdr.ash.u[subset],Bayes=0.1*maxx*findInterval(lfdr.ash.true[subset],seq(0,maxx,length=11))-0.05*maxx,mixfdr = lfdr.mixfdr[subset])

      res[[i]]=rbind(res.lfdr[[i]],res.lfsr[[i]])
    }

    require(reshape2)

    res.melt = melt(res, id.vars=c("Bayes","Scenario","Measure"),variable.name="Method")
    res.melt$Scenario = as.factor(res.melt$Scenario)
    levels(res.melt$Scenario) = c("Scenario 1a","Scenario 1b","Scenario 2")

    cbbPalette <- c("#E69F00","#000000", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
    labels = c('mixfdr','ash (null-biased)','ash (mle)')
    breaks = c("mixfdr","ash.nullbiased","ash.uniform")

    p=ggplot(data=res.melt,aes(Bayes,value,colour=Method)) + 
        facet_grid(Measure ~ Scenario ) +
        #scale_fill_manual(values=cbbPalette) +
        #scale_colour_manual(values=cbbPalette) +
        geom_point(size=1,alpha=0.1) + 
        #stat_smooth(se=FALSE,size=2) +
        stat_summary(fun.data="mean_quant", geom="ribbon", alpha=0.25) +
        geom_abline(colour = "red",size=1) +
        xlab("Truth") +
        ylab("Estimate")

    print(p +scale_y_continuous(limits=c(0,maxy)) +
          scale_x_continuous(limits=c(0,maxx)) +
            scale_colour_manual(values=cbbPalette,breaks=breaks,labels=labels))        
  }

>>>>

```{r}
#Plot pi0 from each method
source("plotpi0.R")

@@

  png("figures/estpi0_sim1sim2.png",height=160,width=540)
    plot_pi0(list(simres1,simres2))
  dev.off()

@@

```{r}
source("plot_lfsr.R")
```

### 2014-02-24 talks/UCstat2014.Rmd
added uc talk

% Shrinkage, False Discovery Rates, and an Alternative to the Zero Assumption
% Matthew Stephens
% 2013/11/1

```{r setup, include=FALSE}
# set global chunk options
opts_chunk$set(cache=TRUE,autodep=TRUE)
dep_auto()
library("qvalue")
library("ashr")
```

@@

- Constrain $g$ to be unimodal about 0.

@@

# FDR - The New Deal

@@

@@

- In practice, implement upper bound by putting prior on $\pi_0$ that encourages it to be big, then estimate $\pi$ by posterior mean (VB).

# Illustration: BRCA data

@@

Compare fitted $f(\beta)$, both estimating $\pi_0$ and fixing $\pi_0=0$.
```{r, echo=FALSE, includ=FALSE}
  hh.ash.fdr = ash(hh.zscore,1,pointmass = TRUE,prior="nullbiased")
  hh.ash.shrink = ash(hh.zscore,1)
```

```{r, echo=FALSE,fig.height=4,fig.cap=""}
  plot(x,cdf.ash(hh.ash.fdr,x),col=2,type="l",lwd=3,ylab="cdf")
  lines(x,cdf.ash(hh.ash.shrink,x),col=3,lwd=3,lty=2)
```

@@

# Identifiability of $\pi_0$: Solution 2

@@

- Suppose you estimate $\Pr(\beta_j<0)=0.98$,  $\Pr(\beta_j>0)=0.01$, $\Pr(\beta_j=0) = 0.01$.

- Should you declare an fdr of 0.01 or 0.02?

- Maybe fsr makes more sense anyway?

# Unification of one-group and two-group solutions

- A possible concern with one-group model: if we assume all $\beta$ are non-zero, but really some are actually zero, might we not underestimate the fsr?

- Consider our example, with $\Pr(\beta_j>0)=0.025$. If we actually allowed
$\beta_j=0$ then possibly all of this probability might actually land at $\beta_j=0$.

- I argue that, assuming symmetry of $g$ near 0, this also provides an upper
bound of how much of the $\Pr(\beta_j<0)=0.975$ might also move to 0.

- Therefore a more conservative estimate of the fsr might be 0.05 (or, more generally, double what you get allowing for point mass)

```{r, echo=FALSE,fig.height=4,fig.cap=""}
  plot(hh.ash.fdr$localfsr,2*hh.ash.shrink$localfsr,xlab="fsr with point mass", ylab="2 * fsr without point mass", xlim=c(0,0.1),ylim=c(0,0.1))
  abline(a=0,b=1,col=2,lwd=2)
```

@@

- So for example we can easily compute fdrs for discoveries other than ``non-zero" (eg compute $\Pr(|\beta_j| > 2 \| \hat\beta_j)$).

@@

```{r, echo=FALSE,fig.height=3,fig.cap=""}
plot(zdat.ash$localfdr,pval,ylab="p value", xlab="ASH local fdr")
```

```{r, echo=FALSE,fig.height=3,fig.cap=""}
plot(zdat.ash$localfdr,pval,ylab="p value", xlab="ASH local fdr")
points(zdat.ash$localfdr[15325],pval[15325],col=2,pch=16)
points(zdat.ash$localfdr[16123],pval[16123],col=2,pch=16)
```

```{r, echo=FALSE}
x = read.table(paste0("../../stat45800/data/nobrega/expression/counts.txt"), header = TRUE)
xx = rowSums(x[,2:5])
x = x[xx>0,]
xx = xx[xx>0]
cbind(x[,1:5],pval,zdat.ash$localfdr)[c(15325,16123),]
```

# Summary: FDR via Adapative Shrinkage

- Both provide a rational approach to identifying ``significant" findings.

- Both are generic and modular: once you have the summary data, you can forget where they came from.

- But by using two numbers ($\hat\beta,s$) instead of one ($p$ values) precision of different measurementscan be better accounted for.

- ASH borrows information for estimation, as well as testing.

@@

- Widely applicable: perhaps anywhere (?) 
where shrinkage is appropriate, requiring only an estimated
effect size and standard error for each object.

- Could also use effect size estimate and $p$ value for each variable, by converting to effect size estimate and (pseudo-) standard error.

@@

@@

@@

- to the several postdoctoral researchers and students
who have worked with me on related topics.

- Especially Mengyin Lu who coded the VB algorithm.

@@

`pandoc -s -S -i --template=my.beamer -t beamer -V theme:CambridgeUS -V colortheme:beaver  ilike-slides.md -o ilike-slides.pdf`

(alternative to produce html slides; but figures would need reworking)
`pandoc -s -S -i -t dzslides --mathjax NSmeet2013.md -o NSmeet2013.html`

@@

>>>>

% False Discovery Rates, A New Deal
% Matthew Stephens
% 2014/2/24

```{r, include=FALSE}
require("qvalue")
require("ashr")
require("ggplot2")
```

```{r setup, include=FALSE}
# set global chunk options
opts_chunk$set(cache=TRUE,autodep=TRUE,warning=FALSE)
dep_auto()
```

# Before we get started: Getting Organized

- Over ~10 years of working with graduate students + postdocs,
I've noticed something.
- Organized researchers get more done (and better!).
- Many of them are more organized than I am!
- Thought: I should get organized; I should help others get organized.

# So what can you do?

- Buy a notebook; bring it to meetings; make notes! 
- Come to meetings with a written agenda. 
- While doing research, record what you did and what the outcome was.
- Use version control [git](http://git-scm.com) and internet repositories ([bitbucket](http://www.bitbucket.org), [github](http://www.github.com)) to organize notes, code, etc.
- Use knitr to help make your research reproducible.

# What are these repository things?

- A repository: a central place in which an aggregation of data is kept and maintained in an organized way (searcharticle.com)
- Great for sharing material across multiple people (eg student and advisor!)
- An amateur example: [http://github.com/stephens999/ash](http://github.com/stephens999/ash)

# What is knitr?

- An R package
- A tool for literate programming
- Text, and R code are interleaved
- When you compile the document, the code is run, and output inserted into the text.
- Great for writing reports, and keeping a track of what you did and what the result was!
- This talk was written with knitr (with RStudio)!

# What is Reproducible Research?

- Principle: when publishing results of computational procedures, we should
publish the code that produced the results.
- "publishing figures or results without the complete software environment could
be compared to a mathematician publishing an announcement of a mathematical theorem without giving the proof" (Buckheit and Donohoe)
- “an article about a computational result is advertising, not scholarship. The actual scholarship is the full software environment, code and data, that produced the result.” [Claerbout]

# Why is reproducibility important?

- Not only because people are forgetful, error-prone, or dishonest!
- Reproducing work is also the first step to extending it.
- Helps communications among researchers (eg student + advisor).
- If you do not publish code implementing your methods, your methods will likely go unused.

# More on git, github, knitr, reproducibility

- Google "The git book", to get started on git.
- Google "Karl Broman github tutorial" for statistics-oriented intro to github.
- Google "donohoe buckheit" for "Wavelab and reproducible research"

@@

- Constrain $g$ to be unimodal about 0; estimate $g$ from data.

@@

# FDR - A New Deal

@@

- Can approximate, arbitrarily closely, any scale mixture of normals.
Includes almost all priors used for sparse regression problems (spike-and-slab, double exponential/Laplace/Bayesian Lasso, horseshoe).

# FDR - A New Deal

@@

- In practice, implement upper bound by using penalized likelihood that encourages $\pi_0$ to be as big as possible.

# Illustration: Simulated Example

```{r,echo=FALSE,include=FALSE}
load("../paper/Rcode/sim1_out.RData") #load simulation results
source("../paper/Rcode/plot_pi0.R")
```

```{r, echo=FALSE,fig.height=4,fig.cap=""}
plot_pi0(list(simres1))
```

@@

```{r, echo=FALSE, include=FALSE}
  hh.ash.fdr = ash(hh.zscore,1,pointmass = TRUE,prior="nullbiased")
  hh.ash.shrink = ash(hh.zscore,1)
```

@@

# Identifiability of $\pi_0$ and the False Sign Rate

@@

# The fsr is more robust than fdr

```{r, echo=FALSE,fig.height=4,fig.cap="",warning=FALSE,message=FALSE}
source("../paper/Rcode/plot_lfsr.R")
  plot_lfsr(list(simres1),0.1,ptype="lfdr")
```

# The fsr is more robust than fdr

```{r, echo=FALSE,fig.height=4,fig.cap="",warning=FALSE,message=FALSE}
  plot_lfsr(list(simres1),0.1,ptype="lfsr")
```

@@

- So for example we can easily compute fdrs for discoveries other than ``non-zero" (eg compute $\Pr(\beta_j > 2 | \hat\beta_j)$).

@@

```{r, fig.height=4,echo=FALSE,fig.cap=""}
plot(zdat[3,],zdat.ash$PosteriorMean,main="Estimates vs Shrunken estimates",xlab="betahat",ylab="shrunk betahat",xlim=c(-2,2),ylim=c(-0.1,.1),prob=T)
abline(a=0,b=1,col=2)
```

# Summary

- ASH provides a generic approach to shrinkage estimation, as well as
false discovery (sign) rates.

- But by using two numbers ($\hat\beta,s$) instead of one ($p$ values or $z$ scores) precision of different measurementscan be better accounted for.

- Unimodal assumption for effects reduces conservatism

- False Sign Rate is more robust to assumptions, and perhaps
therefore preferable, than False Discovery Rate.

@@

- Widely applicable: requiring only an estimated
effect size and standard error for each object.

@@

- Theory anyone?

@@

- Incorporate $t$ likelihood as well as normal.

- Incorporate shrinkage of variances and not just means.

@@

- to the several postdoctoral researchers and students
who have worked with me on related topics.
- Including Scott Powers, Mengyin Lu, Tian Sen, Wei Wang, Zhengrong Xing.

@@

`pandoc -s -S -i --template=my.beamer -t beamer -V theme:CambridgeUS -V colortheme:beaver  slides.md -o slides.pdf`

(alternative to produce html slides; but figures would need reworking)
`pandoc -s -S -i -t dzslides --mathjax slides.md -o slides.html`

@@

- Suppose you estimate $\Pr(\beta_j<0)=0.98$,  $\Pr(\beta_j>0)=0.01$, $\Pr(\beta_j=0) = 0.01$. 
- Should you declare an fdr of 0.01 or 0.02?
- Maybe fsr makes more sense anyway?

Need to fix counts.associate to use fdr method in ash

```{r, echo=FALSE,fig.height=3,fig.cap=""}
plot(zdat.ash$lfdr,pval,ylab="p value", xlab="ASH local fdr")
```

```{r, echo=FALSE,fig.height=3,fig.cap=""}
plot(zdat.ash$lfdr,pval,ylab="p value", xlab="ASH local fdr")
points(zdat.ash$lfdr[15325],pval[15325],col=2,pch=16)
points(zdat.ash$lfdr[16123],pval[16123],col=2,pch=16)
```

```{r, echo=FALSE}
x = read.table(paste0("../../stat45800/data/nobrega/expression/counts.txt"), header = TRUE)
xx = rowSums(x[,2:5])
x = x[xx>0,]
xx = xx[xx>0]
cbind(x[,1:5],pval,zdat.ash$lfdr)[c(15325,16123),]
```

Compare fitted $f(\beta)$, both estimating $\pi_0$ and fixing $\pi_0=0$.
```{r, echo=FALSE,fig.height=4,fig.cap=""}
  x=seq(-4,4,length=100)
  plot(cdf.ash(hh.ash.fdr,x),col=2,type="l",lwd=3,ylab="cdf")
  lines(x,cdf.ash(hh.ash.shrink,x)$y,col=3,lwd=3,lty=2)
```

### 2014-02-24 talks/UCtest.Rmd
added uc talk

- Can approximate, arbitrarily closely, any scale mixture of normals.
Includes almost all priors used for sparse regression problems (spike-and-slab, double exponential/Laplace/Bayesian Lasso, horseshoe).

@@

```{r, echo=FALSE, include=FALSE}
  hh.ash.fdr = ash(hh.zscore,1,pointmass = TRUE,prior="nullbiased")
  hh.ash.shrink = ash(hh.zscore,1)
```

@@

# Identifiability of $\pi_0$ and the False Sign Rate

@@

```{r, echo=FALSE,fig.height=4,fig.cap="",warning=FALSE,message=FALSE}
source("../paper/Rcode/plot_lfsr.R")
  plot_lfsr(list(simres1),0.1,ptype="lfdr")
```

```{r, echo=FALSE,fig.height=4,fig.cap="",warning=FALSE,message=FALSE}
  plot_lfsr(list(simres1),0.1,ptype="lfsr")
```

@@

- So for example we can easily compute fdrs for discoveries other than ``non-zero" (eg compute $\Pr(\beta_j > 2 | \hat\beta_j)$).

- And use it to obtain point estimates and credible intervals for each $\beta_j$, taking account of information from all the other $\beta_j$.

- Because $f(\beta)$ is unimodal, the point estimates will tend to be ``shrunk" towards the overall mean (0).

- Because $f(\beta)$ is estimated from the data, the amount
of shrinkage is adaptive to the data. And because of the role of $s_j$, the amount of shrinkage adapts to the information on each gene.

- So we call the approach ``Adaptive Shrinkage" (ASH).

# Example: ASH applied to mouse data

```{r, fig.height=4,echo=FALSE,fig.cap=""}
hist(zdat[3,],main="Raw effect size estimates",xlab="betahat",xlim=c(-2,2),prob=T,ylim=c(0,3))
```

# Example: ASH applied to mouse data

```{r, fig.height=4,echo=FALSE,fig.cap=""}
hist(zdat[3,],main="Raw effect size estimates",xlab="betahat",xlim=c(-2,2),prob=T,ylim=c(0,3))
t=seq(-2,2,length=201)
lines(t,dnorm(t,0,sd=0.032), col=2, lwd=2)
```

# Example: ASH applied to mouse data

```{r, fig.height=4,echo=FALSE,fig.cap=""}
hist(zdat.ash$PosteriorMean,main="Shrunken estimates",xlab="shrunk betahat",xlim=c(-0.1,0.1),prob=T)
lines(t,dnorm(t,0,sd=0.032), col=2, lwd=2)
```

# Example: ASH applied to mouse data

```{r, fig.height=4,echo=FALSE,fig.cap=""}
plot(zdat[3,],zdat.ash$PosteriorMean,main="Estimates vs Shrunken estimates",xlab="betahat",ylab="shrunk betahat",xlim=c(-2,2),ylim=c(-0.1,.1),prob=T)
abline(a=0,b=1,col=2)
```

# Summary

- ASH provides a generic approach to shrinkage estimation, as well as
false discovery (sign) rates.

- But by using two numbers ($\hat\beta,s$) instead of one ($p$ values or $z$ scores) precision of different measurementscan be better accounted for.

- Unimodal assumption for effects reduces conservatism

- False Sign Rate is more robust to assumptions, and perhaps
therefore preferable, than False Discovery Rate.

# Other Applications

- Widely applicable: requiring only an estimated
effect size and standard error for each object.

- Currently applying it to wavelet shrinkage applications.

# Guarantees?

- ``I think you have some nice ideas. How will you convince
people to use them?" (C Morris)

- Theory anyone?

# Next steps?

- Incorporate $t$ likelihood as well as normal.

- Incorporate shrinkage of variances and not just means.

- Extend to allow $g(\cdot;\pi)$ to depend on covariates $X$.

- Extend to allow for correlations in the measured $\hat\beta_j$.

# Thanks

- to the several postdoctoral researchers and students
who have worked with me on related topics.
- Including Scott Powers, Mengyin Lu, Tian Sen, Wei Wang, Zhengrong Xing.

# Reproducible research

- This document is produced with **knitr**, **Rstudio** and **Pandoc**.

- For more details see my `stephens999/ash` repository at `http://www.github.com/stephens999/ash`

- Website: `http://stephenslab.uchicago.edu`

# Pandoc Command used

`pandoc -s -S -i --template=my.beamer -t beamer -V theme:CambridgeUS -V colortheme:beaver  slides.md -o slides.pdf`

(alternative to produce html slides; but figures would need reworking)
`pandoc -s -S -i -t dzslides --mathjax slides.md -o slides.html`

Here is my session info:

```{r session-info}
print(sessionInfo(), locale=FALSE)
```

# Some odd things in the data

```{r, echo=FALSE,fig.height=3}
  plot(zdat[3,],zdat[4,],ylab="standard error",xlab="beta-hat")
  x[tail(order(zdat[3,])),]
```

# A technicality

- Suppose you estimate $\Pr(\beta_j<0)=0.98$,  $\Pr(\beta_j>0)=0.01$, $\Pr(\beta_j=0) = 0.01$. 
- Should you declare an fdr of 0.01 or 0.02?
- Maybe fsr makes more sense anyway?

# Shrinkage is adaptive to information

Need to fix counts.associate to use fdr method in ash

```{r, echo=FALSE,fig.height=4}
#hist(zdat[3,])
#plot(zdat[3,],zdat.ash$PosteriorMean,xlim=c(-0.2,0.2))
#points(zdat[3,16677],zdat.ash$PosteriorMean[16677],col=2,pch=16)
#points(zdat[3,16079],zdat.ash$PosteriorMean[16079],col=2,pch=16)
#x[16677,]
#x[16079,]
```

```{r, echo=FALSE}
#temp = counts.associate(cc,c(-1,1,-1,1),1)
#temp.tscore = temp$zdat[3,]/temp$zdat[4,]
#temp.pval = ttest.pval(temp.tscore,df=2)
#plot(temp$zdat.ash$localfdr,temp.pval)
#identify(temp$zdat.ash$localfdr,temp.pval)
```

```{r, echo=FALSE,fig.height=3,fig.cap=""}
plot(zdat.ash$lfdr,pval,ylab="p value", xlab="ASH local fdr")
```

# Shrinkage is adaptive to information

```{r, echo=FALSE,fig.height=3,fig.cap=""}
plot(zdat.ash$lfdr,pval,ylab="p value", xlab="ASH local fdr")
points(zdat.ash$lfdr[15325],pval[15325],col=2,pch=16)
points(zdat.ash$lfdr[16123],pval[16123],col=2,pch=16)
```

# Shrinkage is adaptive to information

```{r, echo=FALSE}
x = read.table(paste0("../../stat45800/data/nobrega/expression/counts.txt"), header = TRUE)
xx = rowSums(x[,2:5])
x = x[xx>0,]
xx = xx[xx>0]
cbind(x[,1:5],pval,zdat.ash$lfdr)[c(15325,16123),]
```

# Recall FDR problem 1: q values increased by low count genes
```{r,echo=FALSE, fig.height=4,fig.cap=""}
plot(qval.high$qval,qval$qval[xx>1000],xlab="q values from high-count gene analysis", ylab="q values from all gene analysis", main="q values for high count genes",xlim=c(0,.5),ylim=c(0,.5))
abline(a=0,b=1,col=2)
```

# ASH q values more robust to inclusion of low count genes

```{r, echo=FALSE,fig.height=3,fig.cap=""}
plot(zdat.ash.high$qvalue,zdat.ash$qvalue[xx>1000],xlab="q values from high only",ylab="q values from all data")
abline(a=0,b=1,col=2,lwd=2)
```

Compare fitted $f(\beta)$, both estimating $\pi_0$ and fixing $\pi_0=0$.
```{r, echo=FALSE,fig.height=4,fig.cap=""}
  x=seq(-4,4,length=100)
  plot(cdf.ash(hh.ash.fdr,x),col=2,type="l",lwd=3,ylab="cdf")
  lines(x,cdf.ash(hh.ash.shrink,x)$y,col=3,lwd=3,lty=2)
```

>>>>

- Can approximate, arbitrarily closely, any ``scale mixture of normals" which
includes almost all the priors people hoave used for sparse regression problems (spike-and-slab, ``horseshoe", double exponential/Laplace/Lasso).

@@

Compare fitted $f(\beta)$, both estimating $\pi_0$ and fixing $\pi_0=0$.
```{r, echo=FALSE, include=FALSE}
  hh.ash.fdr = ash(hh.zscore,1,pointmass = TRUE,prior="nullbiased")
  hh.ash.shrink = ash(hh.zscore,1)
```

```{r, echo=FALSE,fig.height=4,fig.cap=""}
  x=seq(-4,4,length=100)
  plot(cdf.ash(hh.ash.fdr,x),col=2,type="l",lwd=3,ylab="cdf")
  lines(x,cdf.ash(hh.ash.shrink,x)$y,col=3,lwd=3,lty=2)
```

@@

# Identifiability of $\pi_0$: Solution 2

@@

- but still need to encourage point mass on zero if want to be ``conservative".

- blah

# Unification of one-group and two-group solutions

- A possible concern with one-group model: if we assume all $\beta$ are non-zero, but really some are actually zero, might we not underestimate the fsr?

- Consider our example, with $\Pr(\beta_j>0)=0.025$. If we actually allowed
$\beta_j=0$ then possibly all of this probability might actually land at $\beta_j=0$.

- I argue that, assuming symmetry of $g$ near 0, this also provides an upper
bound of how much of the $\Pr(\beta_j<0)=0.975$ might also move to 0.

- Therefore a more conservative estimate of the fsr might be 0.05 (or, more generally, double what you get allowing for point mass)

```{r, echo=FALSE,fig.height=4,fig.cap=""}
  plot(hh.ash.fdr$lfsr,2*hh.ash.shrink$lfsr,xlab="fsr with point mass", ylab="2 * fsr without point mass", xlim=c(0,0.1),ylim=c(0,0.1))
  abline(a=0,b=1,col=2,lwd=2)
```

@@

- So for example we can easily compute fdrs for discoveries other than ``non-zero" (eg compute $\Pr(|\beta_j| > 2 | \hat\beta_j)$).
